---
title: "Assignment 5"
author: "Anna Chamot"
date: "2025-03-26"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 5

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
   library(R.utils) # unzip .gz archives
})
# we shouldn't actually need this today, but in case:
# ah <- AnnotationHub(localHub=TRUE)
# ensdb <- ah[["AH89211"]]
```

## Using peaks from ENCODE, identify bivalent domains (H3K27me3 + H3K4me3) from (mouse or human) embryonic stem cells (ESC)


  * H3K27me3
  
```{r}
dir.create("peaks")

download.file("https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz", "peaks/H3K27me3.bed.gz") 
gunzip("peaks/H3K27me3.bed.gz", remove=FALSE)

peaks_H3K27me3 <- rtracklayer::import("peaks/H3K27me3.bed", format="narrowPeak")
print(length(peaks_H3K27me3))
head(peaks_H3K27me3)
``` 

  * H3K4me3

```{r}
dir.create("peaks")

download.file("https://www.encodeproject.org/files/ENCFF974BMC/@@download/ENCFF974BMC.bed.gz", "peaks/H3K4me3.bed.gz") 
gunzip("peaks/H3K4me3.bed.gz", remove=FALSE)

peaks_H3K4me3 <- rtracklayer::import("peaks/H3K4me3.bed", format="narrowPeak")
print(length(peaks_H3K4me3))
head(peaks_H3K4me3)
``` 

```{r}
# Find the overlapping peaks 
hits <- findOverlaps(peaks_H3K27me3, peaks_H3K4me3)
# Find the genomic regions of the overlapping peaks
bivalent_domains <- pintersect(
  peaks_H3K27me3[queryHits(hits)],
  peaks_H3K4me3[subjectHits(hits)]
)
print(length(hits))
head(bivalent_domains)
```

## Question: what happens to those regions upon differentiation?

### Choose one differentiated cell type (e.g. erythroblasts, fibroblast, B cells, etc.)

### Download the H3K27me3 and H3K4me3 peaks from this cell type

  * H3K27me3
  
```{r}
dir.create("peaks")

download.file("https://www.encodeproject.org/files/ENCFF728FZO/@@download/ENCFF728FZO.bed.gz", "peaks/H3K27me3_erythroblasts.bed.gz") 
gunzip("peaks/H3K27me3_erythroblasts.bed.gz", remove=FALSE)

peaks_H3K27me3_erythroblasts <- rtracklayer::import("peaks/H3K27me3_erythroblasts.bed", format="narrowPeak")
print(length(peaks_H3K27me3_erythroblasts))
head(peaks_H3K27me3_erythroblasts)
``` 

  * H3K4me3

```{r}
dir.create("peaks")

download.file("https://www.encodeproject.org/files/ENCFF339CXH/@@download/ENCFF339CXH.bed.gz", "peaks/H3K4me3_erythroblasts.bed.gz") 
gunzip("peaks/H3K4me3_erythroblasts.bed.gz", remove=FALSE)

peaks_H3K4me3_erythroblasts <- rtracklayer::import("peaks/H3K4me3_erythroblasts.bed", format="narrowPeak")
print(length(peaks_H3K4me3_erythroblasts))
head(peaks_H3K4me3_erythroblasts)
``` 


### How many of the mESC bivalent domains are, in this differentiated cell type, overlapping either mark or their combination (in this differentiated cell type)?


```{r}
# Overlap of bivalent domains with H3K27me3 in erythroblasts
overlap_H3K27me3 <- queryHits(findOverlaps(bivalent_domains, peaks_H3K27me3_erythroblasts))
length(unique(overlap_H3K27me3))  # Number of unique bivalent domains retaining H3K27me3

# Overlap of bivalent domains with H3K4me3 in erythroblasts
overlap_H3K4me3 <- queryHits(findOverlaps(bivalent_domains, peaks_H3K4me3_erythroblasts))
length(unique(overlap_H3K4me3))  # Number of unique bivalent domains retaining H3K4me3

# Still bivalent in erythroblasts
bivalent_erythroblasts <- intersect(overlap_H3K27me3, overlap_H3K4me3)
length(bivalent_erythroblasts)

```

